---
title: "Country-level temperature trends"
output:
  rmarkdown::html_vignette:
    self_contained: true
vignette: >
  %\VignetteIndexEntry{Country-level temperature trends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#",
  eval = TRUE,
  fig.width = 10,
  fig.height = 7,
  fig.dpi = 300,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)
```

```{r setup}
library(varunayan)
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
```

`get_era5_country_temperature()` resolves a country boundary, downloads
monthly ERA5 data, and returns cosine-latitude area-weighted national
averages. One call replaces the manual workflow of finding boundaries,
writing GeoJSON, calling `era5ify_geojson()`, and aggregating grid cells.

Monthly mean, max, and min temperatures for four European and four South Asian
countries, 2024.

## Download data

The `variables` argument accepts shorthands `"mean"`, `"max"`, and `"min"`.

```{r download, eval = FALSE}
countries <- c(
  "France", "Germany", "Spain", "Norway",
  "India", "Pakistan", "Bangladesh", "Nepal"
)

all_data <- lapply(countries, function(cntry) {
  df <- get_era5_country_temperature(
    country    = cntry,
    start_date = "2024-01-01",
    end_date   = "2024-12-31",
    variables  = c("mean", "max", "min")
  )
  df$country <- cntry
  df
})

temps <- bind_rows(all_data)
```

```{r load-data, echo = FALSE}
temps <- readRDS(system.file("extdata", "country_temps.rds", package = "varunayan"))
```

```{r preview}
head(temps)
```

Tag each country with its region for faceting.

```{r regions}
region_map <- c(
  France = "Europe", Germany = "Europe", Spain = "Europe", Norway = "Europe",
  India = "South Asia", Pakistan = "South Asia",
  Bangladesh = "South Asia", Nepal = "South Asia"
)

temps$region <- region_map[temps$country]
temps$month_label <- factor(month.abb[temps$month], levels = month.abb)

# Cool tones for Europe, warm tones for South Asia
country_colors <- c(
  France = "#2166AC", Germany = "#4393C3",
  Spain = "#92C5DE", Norway = "#053061",
  India = "#B2182B", Pakistan = "#D6604D",
  Bangladesh = "#F4A582", Nepal = "#67001F"
)
```

## Seasonal heatmap

```{r heatmap, fig.height = 6}
ggplot(temps, aes(x = month_label, y = country, fill = temperature_mean)) +
  geom_tile(color = "white", linewidth = 0.4) +
  facet_grid(region ~ ., scales = "free_y", space = "free_y") +
  scale_fill_distiller(
    palette = "RdYlBu", direction = -1,
    name = "Mean temp (\u00B0C)"
  ) +
  labs(
    x = NULL, y = NULL,
    title = "Mean monthly temperature by country (2024)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    strip.text.y = element_text(angle = 0, face = "bold")
  )
```

South Asian countries stay above 20 &deg;C year-round. European countries
swing 20+ &deg;C between winter and summer, with Norway dropping below
freezing from November through March.

## Monthly trend lines with min/max ribbon

```{r trendlines, fig.height = 8}
ggplot(temps, aes(x = month, color = country, fill = country)) +
  geom_ribbon(aes(ymin = temperature_min, ymax = temperature_max),
    alpha = 0.15, color = NA
  ) +
  geom_line(aes(y = temperature_mean), linewidth = 0.9) +
  geom_point(aes(y = temperature_mean), size = 1.5) +
  facet_wrap(~region, ncol = 1, scales = "free_y") +
  scale_color_manual(values = country_colors) +
  scale_fill_manual(values = country_colors) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(
    x = NULL, y = "Temperature (\u00B0C)",
    title = "Monthly temperature with min-max range (2024)",
    subtitle = "Ribbon: national-average daily min to daily max",
    color = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Diurnal range

The gap between daily max and min (DTR) is wider in continental and arid
climates than in humid tropical ones.

```{r dtr, fig.height = 6}
temps$dtr <- temps$temperature_max - temps$temperature_min

ggplot(temps, aes(x = month, y = dtr, color = country)) +
  geom_line(linewidth = 0.9) +
  geom_point(size = 1.5) +
  facet_wrap(~region, ncol = 1) +
  scale_color_manual(values = country_colors) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(
    x = NULL, y = "Diurnal temperature range (\u00B0C)",
    title = "Monthly diurnal temperature range (2024)",
    color = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

## Regional comparison

All eight countries on a single axis, with line type separating the two regions.

```{r comparison, fig.height = 6}
ggplot(temps, aes(
  x = month, y = temperature_mean,
  color = country, linetype = region
)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  scale_color_manual(values = country_colors) +
  scale_linetype_manual(values = c("Europe" = "solid", "South Asia" = "dashed")) +
  labs(
    x = NULL, y = "Mean temperature (\u00B0C)",
    title = "Europe vs South Asia: monthly temperature (2024)",
    color = NULL, linetype = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```
