---
title: "CRU TS vs ERA5: India temperature comparison (2000-2018)"
output:
  rmarkdown::html_vignette:
    self_contained: true
vignette: >
  %\VignetteIndexEntry{CRU TS vs ERA5: India temperature comparison (2000-2018)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
has_data <- nzchar(system.file("extdata", "india_era5_cru_comparison.rds",
  package = "varunayan"))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#",
  eval = has_data,
  fig.width = 8,
  fig.height = 4,
  fig.dpi = 150,
  warning = FALSE,
  message = FALSE,
  out.width = "100%"
)
```

```{r setup}
library(varunayan)
library(dplyr)
library(ggplot2)
library(tidyr)
library(patchwork)
```

How well do [ERA5](https://cds.climate.copernicus.eu/datasets/reanalysis-era5-single-levels?tab=overview) and [CRU TS](https://crudata.uea.ac.uk/cru/data/hrg/) agree on monthly temperature over India? They are two independent records covering 2000-2018:

- **ERA5** - ECMWF reanalysis at 0.25°, pulled with `era5ify_bbox`
- **CRU TS v4.07** - station-interpolated observations from UEA at 0.5°, pulled with `cru_ts_bbox`

CRU TS variable dictionary with ERA5 equivalents:

```{r variable_dict}
list_cru_ts_variables() |>
  select(variable, name, unit, era5_equivalent)
```

## Download

```{r download, eval = FALSE}
# ERA5 monthly 2m_temperature for India
era5_tmp <- era5ify_bbox(
  request_id = "india_era5_tmp_monthly",
  variables = "2m_temperature",
  start_date = "2000-01-01", end_date = "2018-12-31",
  north = 37, south = 6, east = 98, west = 68,
  frequency = "monthly", resolution = 0.5
)

# CRU TS monthly mean temperature for India
cru_tmp <- cru_ts_bbox(
  variable = "tmp",
  start_year = 2000, end_year = 2018,
  north = 37, south = 6, east = 98, west = 68
)
```

```{r load_data}
saved <- readRDS(system.file("extdata", "india_era5_cru_comparison.rds",
  package = "varunayan"
))
era5_tmp <- saved$era5_tmp
cru_tmp <- saved$cru_tmp
```

## Monthly time series

Spatial mean temperature across all India grid cells, month by month:

```{r monthly_ts, fig.height = 4}
era5_monthly <- era5_tmp |>
  group_by(year, month) |>
  summarise(temp = mean(value, na.rm = TRUE), .groups = "drop") |>
  mutate(source = "ERA5", date = as.Date(sprintf("%d-%02d-01", year, month)))

cru_monthly <- cru_tmp |>
  group_by(year, month) |>
  summarise(temp = mean(value, na.rm = TRUE), .groups = "drop") |>
  mutate(source = "CRU TS", date = as.Date(sprintf("%d-%02d-01", year, month)))

temp_all <- bind_rows(era5_monthly, cru_monthly)

ggplot(temp_all, aes(x = date, y = temp, color = source)) +
  geom_line(alpha = 0.5, linewidth = 0.5) +
  geom_smooth(method = "loess", span = 0.07, se = FALSE, linewidth = 1.1) +
  scale_x_date(date_breaks = "2 years", date_labels = "%Y") +
  scale_color_manual(values = c("ERA5" = "#E74C3C", "CRU TS" = "#2E86AB")) +
  labs(
    x = NULL, y = "Mean temperature (\u00b0C)", color = NULL,
    title = "India: Monthly Mean Temperature — ERA5 vs CRU TS (2000-2018)",
    subtitle = "Spatial mean across all grid cells"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

## Comparison

```{r scatter, fig.width = 5, fig.height = 5}
temp_wide <- inner_join(
  era5_monthly |> select(year, month, ERA5 = temp),
  cru_monthly |> select(year, month, `CRU TS` = temp),
  by = c("year", "month")
)

cor_val <- cor(temp_wide$ERA5, temp_wide$`CRU TS`, use = "complete.obs")
bias <- mean(temp_wide$ERA5 - temp_wide$`CRU TS`, na.rm = TRUE)

ggplot(temp_wide, aes(x = `CRU TS`, y = ERA5)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "grey40") +
  geom_point(alpha = 0.5, size = 1.8, color = "#34495E") +
  geom_smooth(
    method = "lm", se = TRUE,
    color = "#E74C3C", fill = "#E74C3C", alpha = 0.15
  ) +
  coord_fixed() +
  labs(
    x = "CRU TS (\u00b0C)", y = "ERA5 (\u00b0C)",
    title = "ERA5 vs CRU TS: Spatial Mean Monthly Temperature",
    subtitle = sprintf("r = %.3f   bias (ERA5 \u2212 CRU TS) = %+.2f\u00b0C", cor_val, bias)
  ) +
  theme_minimal(base_size = 11)
```

## Seasonal cycle

Averaged over 2000-2018, the seasonal cycles nearly overlap:

```{r seasonal, fig.height = 4}
seasonal <- temp_all |>
  group_by(source, month) |>
  summarise(temp = mean(temp, na.rm = TRUE), .groups = "drop") |>
  mutate(month_label = factor(month.abb[month], levels = month.abb))

ggplot(seasonal, aes(x = month_label, y = temp, color = source, group = source)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 2.5) +
  scale_color_manual(values = c("ERA5" = "#E74C3C", "CRU TS" = "#2E86AB")) +
  labs(
    x = NULL, y = "Mean temperature (\u00b0C)", color = NULL,
    title = "India: Seasonal Temperature Cycle (2000-2018 mean)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

## Spatial bias map

To compare grid cell by grid cell, we snap ERA5 (0.25°) onto the CRU TS 0.5° grid and look at the difference:

```{r spatial_bias, fig.width = 10, fig.height = 5}
# Snap ERA5 (0.25° grid at whole degrees) to CRU TS grid (0.5° grid offset by 0.25°).
# CRU TS grid centers: floor(x/0.5)*0.5 + 0.25 → 6.25, 6.75, 7.25, ...
era5_spatial <- era5_tmp |>
  mutate(
    lat_bin = floor(latitude / 0.5) * 0.5 + 0.25,
    lon_bin = floor(longitude / 0.5) * 0.5 + 0.25
  ) |>
  group_by(lat_bin, lon_bin) |>
  summarise(era5_mean = mean(value, na.rm = TRUE), .groups = "drop")

cru_spatial <- cru_tmp |>
  group_by(latitude, longitude) |>
  summarise(cru_mean = mean(value, na.rm = TRUE), .groups = "drop")

spatial_combined <- inner_join(
  era5_spatial, cru_spatial,
  by = c("lat_bin" = "latitude", "lon_bin" = "longitude")
) |>
  mutate(bias = era5_mean - cru_mean)

p_era5 <- ggplot(spatial_combined, aes(x = lon_bin, y = lat_bin, fill = era5_mean)) +
  geom_tile() +
  scale_fill_distiller(
    palette = "RdYlBu", direction = -1,
    name = "\u00b0C", limits = c(10, 36)
  ) +
  coord_fixed() +
  labs(title = "ERA5 Annual Mean") +
  theme_void(base_size = 10) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p_cru <- ggplot(spatial_combined, aes(x = lon_bin, y = lat_bin, fill = cru_mean)) +
  geom_tile() +
  scale_fill_distiller(
    palette = "RdYlBu", direction = -1,
    name = "\u00b0C", limits = c(10, 36)
  ) +
  coord_fixed() +
  labs(title = "CRU TS Annual Mean") +
  theme_void(base_size = 10) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

bias_lim <- max(abs(spatial_combined$bias), na.rm = TRUE)
p_bias <- ggplot(spatial_combined, aes(x = lon_bin, y = lat_bin, fill = bias)) +
  geom_tile() +
  scale_fill_gradient2(
    low = "#2166AC", mid = "white", high = "#B2182B",
    midpoint = 0, limits = c(-bias_lim, bias_lim),
    name = "\u00b0C"
  ) +
  coord_fixed() +
  labs(title = "ERA5 \u2212 CRU TS") +
  theme_void(base_size = 10) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))

p_era5 + p_cru + p_bias + plot_layout(ncol = 3, guides = "collect") &
  theme(legend.position = "bottom")
```

## Annual trend

```{r annual_trend, fig.height = 4}
annual <- temp_all |>
  group_by(source, year) |>
  summarise(temp = mean(temp, na.rm = TRUE), .groups = "drop")

ggplot(annual, aes(x = year, y = temp, color = source)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.5) +
  geom_smooth(method = "lm", se = FALSE, linetype = "dashed", linewidth = 0.8) +
  scale_color_manual(values = c("ERA5" = "#E74C3C", "CRU TS" = "#2E86AB")) +
  labs(
    x = NULL, y = "Annual mean temperature (\u00b0C)", color = NULL,
    title = "India: Annual Temperature Trend — ERA5 vs CRU TS",
    subtitle = "Dashed = linear trend"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top", panel.grid.minor = element_blank())
```

## When to use which

```{r comparison_table, echo = FALSE}
comparison <- data.frame(
  Feature = c(
    "Spatial resolution", "Temporal coverage", "Temporal resolution",
    "Data source", "Credentials required", "Key variables"
  ),
  `CRU TS v4.07` = c(
    "0.5\u00b0", "1901-2022", "Monthly",
    "Station interpolation (land only)", "No (HTTP download)",
    "tmp, tmx, tmn, pre, vap, cld, wet, frs, dtr, pet"
  ),
  ERA5 = c(
    "0.25\u00b0 (or coarser)", "1940-present", "Hourly, daily, monthly",
    "Reanalysis (model + obs, land + ocean)", "Yes (CDS account)",
    "100+ variables incl. wind, radiation, soil moisture"
  ),
  check.names = FALSE,
  stringsAsFactors = FALSE
)
knitr::kable(comparison, align = "lll")
```

CRU TS is the simpler choice when you need long historical baselines without CDS credentials, or monthly land-only temperature is enough. ERA5 wins when you need sub-monthly frequency, ocean coverage, or variables beyond temperature.
